---
title: "Find and map pair sites"
author: "emlombardi"
date: "2025-01-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(giscoR)
library(leaflet)

```

## Load Data

```{r} 
dat <- read.delim("~/Desktop/Research/UMD/Herbarium Research/Collection Distances Calculator/Longitudinal Collections Location ID /Bstricta_GBIF_Jan32025.csv")

names(dat)
table(dat$basisOfRecord)

#conver dat to spatial dataframe
dat.sf <- st_as_sf(dat, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326) %>% st_transform(crs = 3857)
```

## Split and Subset Data
Create subsets of data based on collection type and timing. For now, I'll simply split so that we have a set of herbarium specimens and a set of iNaturalist observations. 

```{r}
#herbarium data
herb <- dat.sf %>%
  filter(basisOfRecord == "PRESERVED_SPECIMEN") %>%
  select(gbifID, scientificName, locality, stateProvince, elevation, year)

#observation data
obs <- dat.sf %>%
  filter(basisOfRecord == "HUMAN_OBSERVATION")%>%
  select(gbifID, scientificName, locality, stateProvince, elevation, year)
```

## Buffer around herbarium collection locations
Create a 1 kilometer buffer around each record in the herbarium dataframe. 

```{r}
herb.buffer <- st_buffer(herb, dist = 1000)

#find observations within the 1 kilometer buffer
match <- st_within(obs, herb.buffer, sparse = FALSE)

# Create a dataframe with matches
matched <- tibble(
  herbarium_id = rep(herb$gbifID, each = nrow(obs)),
  obs_id = rep(obs$gbifID, times = nrow(herb)),
  within_1km = as.vector(match)
) %>%
  filter(within_1km) # Keep only the matches

# Join back with original data for additional details if needed
matched <- matched %>%
  left_join(herb, by = c("herbarium_id" = "gbifID")) %>%
  left_join(select(obs, gbifID, year), by = c("obs_id" = "gbifID"))

matched <- matched %>%
  mutate(yearDiff = abs(year.x - year.y))

```

## Visualize
For each of the herbarium specimens that have a human observation made within a kilometer, what is the general time difference?
```{r}
hist(matched$yearDiff)
summary(matched$yearDiff)

#Hmmm...seems like we have clusters that we might want to contend with
length(unique(matched$herbarium_id)) 
length(unique(matched$obs_id))

```

### Map

```{r}
matched %>%
  ggplot() +
  geom_sf(aes(geometry = geometry.x), color = "red", size = 2, alpha = 0.7) +
  geom_sf(aes(geometry = geometry.y), color = "black", size = .5, alpha = 0.7) +
  ggtitle("Herbarium and Observation Points") +
  theme_minimal()

```
### interactive mapping
```{r}
# Load required packages
library(leaflet)
library(sf)
library(dplyr)

# Assuming 'herb' and 'obs' are already sf objects (spatial data frames)
# Ensure that both datasets have the same CRS, and reproject if necessary
herb <- dat.sf %>%
  filter(stateProvince == "Colorado") %>% #subsetting this just so there are fewer data for now
  filter(basisOfRecord =="PRESERVED_SPECIMEN") %>%
  st_transform(crs = 4326)  
obs <- dat.sf %>%
  filter(stateProvince == "Colorado") %>%
  filter(basisOfRecord =="HUMAN_OBSERVATION") %>%
  st_transform(crs = 4326)  


# Create a 1-kilometer buffer around 'herb' points
herb_buffer <- st_buffer(herb, dist = 1000)  # Buffer distance in meters
plot(herb_buffer$geometry)


# Create leaflet map
map <- leaflet() %>%
  addTiles() %>%
  
  # Add semi-transparent buffer around herb records
  addPolygons(
    data = herb_buffer,
    color = "salmon",       # Border color
    fillColor = "salmon",   # Fill color
    fillOpacity = 0.3,    # Transparency of the buffer
    weight = 1,
    popup = "Herbarium Buffer"
  ) %>%
  
  # Add 'herb' points in red
  addCircleMarkers(
    data = herb,
    color = "salmon",
    radius = 2,
    fillOpacity = 0.7,
    weight = 1,
    popup =  ~paste("Herbarium Collections, ID: ", gbifID)
  ) %>%
  
  # Add 'obs' points in black
  addCircleMarkers(
    data = obs,
    color = "slategrey",
    radius = 5,
    fillOpacity = 0.7,
    weight = 2,
    popup = ~paste("Observation, ID: ", gbifID)
  ) %>%
  
  # Add a legend for point types
  addLegend(
    position = "bottomright",
    colors = c("salmon", "slategrey"),
    labels = c("Herbarium Collections", "Human Observations"),
    title = "Record Type"
  )

# Display the map
map
```

### subset map
include only the observations that were within the 1 kilometer buffer 
```{r}
# subset the observations
obs.near <- st_intersection(obs, herb_buffer)

map2 <- leaflet() %>%
  addTiles() %>%
  
  # Add semi-transparent buffer around herb records
  addPolygons(
    data = herb_buffer,
    color = "salmon",       # Border color
    fillColor = "salmon",   # Fill color
    fillOpacity = 0.3,      # Transparency of the buffer
    weight = 1,
    popup = "Herbarium Buffer"
  ) %>%
  
  # Add 'herb' points in red
  addCircleMarkers(
    data = herb,
    color = "salmon",
    radius = 2,
    fillOpacity = 0.7,
    weight = 1,
    popup = ~paste("Herbarium Collections, ID: ", gbifID)
  ) %>%
  
  # Add 'obs' points in black, but only those within the buffer
  addCircleMarkers(
    data = obs.near,
    color = "goldenrod",
    radius = 3,
    fillOpacity = 0.7,
    weight = 2,
    popup = ~paste("Observation, ID: ", gbifID)
  ) %>%
  
  # Add a legend for point types
  addLegend(
    position = "bottomright",
    colors = c("salmon", "goldenrod"),
    labels = c("Herbarium Collections", "Human Observations within Buffer"),
    title = "Record Type"
  )

map2


```
